---
title: "working_paper_outputs"
author: "Todd Nobles"
format: docx
---

Read in our working dataset created in data_prep.qmd

```{r}
rm(list = ls())
library(tidyverse)

se_sample <- readRDS("se_sample.rds")
se_sample.5 <- readRDS("se_sample_5.rds")
se_sample_cleanest <- readRDS("se_sample_cleanest.rds") 
se_sample.5_cleanest <- readRDS("se_sample_5_cleanest.rds")
```


```{r}
se_sample <- se_sample %>% 
  mutate(y1_status_v2 = factor(y1_status_v2, labels = c("W&S", "Self-Employed", "Unemployed")))


person_level <- se_sample %>%
  arrange(ssuid_spanel_pnum, swave, monthcode) %>% 
  # getting to person-year level
  group_by(ssuid_spanel_pnum, swave) %>%
  summarize(# the following three are all constant within individuals so we just grab their first value
    y1_status_v2 = first(y1_status_v2),
    race_eth_fct_cpsd = first(race_eth_fct_cpsd),
    sex_fct = first(sex_fct),
    # here we take their highest education value ever observed 
    educ_fct = last(educ_fct),
    tpearn_annual = sum(tpearn),
    industry = first(industry_fct)) %>% 
  # summarizing to person-level 
  group_by(ssuid_spanel_pnum) %>% 
  summarize(
    avg_earnings = mean(tpearn_annual),
    median_earnings = quantile(tpearn_annual, .5),
    y1_status_v2 = first(y1_status_v2),
    race_eth_fct_cpsd = first(race_eth_fct_cpsd),
    sex_fct = first(sex_fct),
    # here we take their highest education value ever observed 
    educ_fct = last(educ_fct),
    industry = first(industry)
  ) %>%
  ungroup()



```


```{r}
se_sample_cleanest <- se_sample_cleanest %>% 
  mutate(y1_status_v2 = fct_recode(as_factor(y1_status_v2), "Self-Emp" = "2", "Unemployed"="4", "W&S" = "1"))


person_level_clean <- se_sample_cleanest %>%
  arrange(ssuid_spanel_pnum, swave, monthcode) %>% 
  # getting to person-year level
  group_by(ssuid_spanel_pnum, swave) %>%
  summarize(# the following three are all constant within individuals so we just grab their first value
    y1_status_v2 = first(y1_status_v2),
    race_eth_fct_cpsd = first(race_eth_fct_cpsd),
    sex_fct = first(sex_fct),
    # here we take their highest education value ever observed 
    educ_fct = last(educ_fct),
    tpearn_annual = sum(tpearn),
    industry = first(industry_fct)) %>% 
  # summarizing to person-level 
  group_by(ssuid_spanel_pnum) %>% 
  summarize(
    avg_earnings = mean(tpearn_annual),
    median_earnings = quantile(tpearn_annual, .5),
    y1_status_v2 = first(y1_status_v2),
    race_eth_fct_cpsd = first(race_eth_fct_cpsd),
    sex_fct = first(sex_fct),
    # here we take their highest education value ever observed 
    educ_fct = last(educ_fct),
    industry = first(industry)
  ) %>%
  ungroup()


```

```{r, setting labels}
# library(labelled)
# person_level <- person_level %>%
#   set_variable_labels(
#     average_earnings = "Age (years)",
#     sex = "Gender",
#     bmi = "Body Mass Index (kg/m²)"
#   )
```


## Table 1 
Descriptive statistics for self-employed sample by initial employment status
Person-level table
```{r}
library(gtsummary)
person_level %>%
  select(avg_earnings, race_eth_fct_cpsd, y1_status_v2, educ_fct, sex_fct, industry) %>% 
  tbl_summary(by = y1_status_v2,
             statistic = list(all_continuous() ~ "{mean} ({sd})")) %>% 
  add_overall()
```


```{r}
library(modelsummary)
datasummary(avg_earnings*educ_fct ~ mean * y1_status_v2,
            data = person_level,
            fmt = fmt_sprintf("%.0f"))

datasummary(avg_earnings*race_eth_fct_cpsd ~ mean * y1_status_v2,
            data = person_level,
            fmt = fmt_sprintf("%.0f"))

datasummary(avg_earnings*sex_fct ~ mean * y1_status_v2,
            data = person_level,
            fmt = fmt_sprintf("%.0f"))
```

```{r}
datasummary(avg_earnings*educ_fct ~ mean * y1_status_v2,
            data = person_level_clean,
            fmt = fmt_sprintf("%.0f"))

datasummary(avg_earnings*race_eth_fct_cpsd ~ mean * y1_status_v2,
            data = person_level_clean,
            fmt = fmt_sprintf("%.0f"))

datasummary(avg_earnings*sex_fct ~ mean * y1_status_v2,
            data = person_level_clean,
            fmt = fmt_sprintf("%.0f"))
```


To-dos: 
- look at annual earnings 
- test descriptives with weights included 
- look at regressions, 
- test regressions with weights included


```{r}

```








# Regressions for earnings
Using person-year level data

https://cran.r-project.org/web/packages/robustlmm/vignettes/rlmer.pdf 
clubSandwich


```{r function to log transform vars with negative values in them}
# feed it a vector to convert to log transform that accounts for negative values 
log_transform <- function(x) {
  min_x <- min(x)*-1
  log(x + min_x + 1)
}
```



```{r}

## getting down to a dataset that thas their demographci vars and yearly earnings at the person-year level
temp <- se_sample %>% 
  filter(monthcode %% 12 == 0) %>% 
  select(ssuid_spanel_pnum, 
         y1_status_v2, 
         unemp_f12,
         educ_fct, 
         race_eth_fct_cpsd,
         sex_fct,
         immig_fct,
         parent_std,
         industry_fct,
         tage,
         calyear,
         prftb, 
         wpfinwgt,
         val) %>% 
  mutate(unemp_f12_6 = factor(ifelse(unemp_f12 >=6, 1,0), labels = c("Not Unemployed", "Unemployed")),
         industry_fct = as_factor(industry_fct),
         industry_num = as.numeric(industry_fct))

temp_earnings <- se_sample %>% 
  group_by(ssuid_spanel_pnum, calyear) %>% 
  summarize(tpearn_annual= sum(tpearn)) %>% 
  ungroup()


person_year <- left_join(temp_earnings, temp, by = c("ssuid_spanel_pnum", "calyear")) %>% 
  mutate(ln_tpearn_annual = log_transform(tpearn_annual))
```

```{r}
library(lme4)

m1_unemp <- lmer(ln_tpearn_annual ~ 
                   unemp_f12_6  + 
                   educ_fct + 
                   tage + 
                   I(tage^2) + 
                   sex_fct + 
                   immig_fct + 
                   calyear + 
                   industry_fct + 
                   (1 | ssuid_spanel_pnum), 
                 data = person_year)

m2_unemp <- update(m1_unemp, . ~ .  + race_eth_fct_cpsd)
m3_unemp <- update(m2_unemp, . ~ . + race_eth_fct_cpsd:unemp_f12_6)
m4_unemp <- update(m2_unemp, . ~ . + sex_fct:unemp_f12_6)



#A more precise p-value can be computed using p.val = "kr"
# regressions for full sample with y1_status_v2
```


```{r}
# options for vcov.fun values https://jepusto.github.io/clubSandwich/reference/vcovCR.html

sjPlot::tab_model( m1_unemp,
                   m2_unemp,
                   m3_unemp,
                   m4_unemp,
                  collapse.ci = TRUE,  
                  p.style = "stars", 
                  prefix.labels = "label", 
                  vcov.args = list(cluster = person_year$ssuid_spanel_pnum), vcov.fun = "CR1" )
```


```{r}
sjPlot::tab_model(m1_unemp,
                   m2_unemp,
                   m3_unemp,
                   m4_unemp,
                  collapse.ci = TRUE,  
                  p.style = "stars", 
                  prefix.labels = "label")
```


```{r}
# regressions for se sample with y1_status_v2
m1_status <- lmer(ln_tpearn_annual ~ 
                   y1_status_v2  + 
                   educ_fct + 
                   tage + 
                   I(tage^2) + 
                   sex_fct + 
                   immig_fct + 
                   calyear + 
                   industry_fct + 
                   (1 | ssuid_spanel_pnum), 
                 data = person_year)

m2_status <- update(m1_status, . ~ .  + race_eth_fct_cpsd)
m3_status <- update(m2_status, . ~ . + race_eth_fct_cpsd:y1_status_v2)
m4_status <- update(m2_status, . ~ . + sex_fct:y1_status_v2)



```

```{r}
sjPlot::tab_model(m1_status,
                   m2_status,
                   m3_status,
                   m4_status,
                  collapse.ci = TRUE,  
                  p.style = "stars", 
                  prefix.labels = "label")
```











# Profits

```{r}
# models for profit


# models for business value 

```



```{r}
# profit for profit positive
# profit for profit10k

#ln_tjb_prftb 
# ln_tbsjval
```

