---
title: "initial_data_prep"
format: html
---
 
 
This is the first script in the process for recreating the Stata anaylses for the SIPP data. This script reads in the raw SIPP files, extracts relevant work/income variables, reshapes the person-month-job variables to align to the person-month variables. 


```{r}
library(data.table)

ds <- c("PU2014W1", "PU2014W2", "PU2014W3", "PU2014W4", "pu2018", "pu2019", "pu2020", "pu2021")

for (x in ds) {
  cols <- colnames(fread(paste0("/Volumes/ExtremeSSD/SIPP Data Files/delims/",x, ".csv"), sep = "|", nrows = 0))
  newname <- paste0(tolower(x))
  print(newname)
  
  
  
  # handling annoying case inconsistencies in their coding.
  target_columns <- c('SSUID', 'PNUM', 'MONTHCODE', 'ERESIDENCEID', 'ERELRPE', 'SPANEL', 
                      'SWAVE', 'SHHADID', 'WPFINWGT', 'ESEX', 'TAGE', 'TAGE_EHC', 'ERACE', 
                      'EORIGIN', 'EEDUC', 'EMS', 'TCEB', 'EBORNUS', 'TDEBT_AST', 'THDEBT_AST', 
                      'TOEDDEBTVAL', 'TDEBT_CC', 'THDEBT_CC', 'TDEBT_ED', 'THDEBT_ED', 
                      'TDEBT_HOME', 'THDEBT_HOME', 'TDEBT_BUS', 'TTHR401VAL', 'TIRAKEOVAL', 
                      'TVAL_AST', 'THVAL_AST', 'TNETWORTH', 'THNETWORTH', 'TVAL_HOME', 
                      'THVAL_HOME', 'TEQ_HOME', 'THEQ_HOME', 'TEQ_BUS', 'TPTOTINC', 
                      'TPEARN', 'ENJFLAG')
  
  # Match column names after coercing all to same case
  selected_columns <- cols[tolower(cols) %in% tolower(target_columns)]
 
  pu <- fread(paste0("/Volumes/ExtremeSSD/SIPP Data Files/delims/",x, ".csv"), 
              sep = "|", 
              select = c(selected_columns, 
                         grep("EJB\\d_JBORSE", cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_CLWRK", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_EMPB", cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_INCPB", cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_JOBID", cols, value = TRUE, ignore.case = TRUE),
                         grep("AJB\\d_JOBID", cols, value = TRUE, ignore.case = TRUE)
                         grep("EJB\\d_STARTWK", cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_ENDWK", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_MWKHRS", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_IND", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_OCC",cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_TYPPAY1", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_GAMT1", cols, value = TRUE, ignore.case = TRUE),
                         grep("EJB\\d_BSLRYB",cols, value = TRUE, ignore.case = TRUE),
                         grep("TBSJ\\dVAL", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_PRFTB", cols, value = TRUE, ignore.case = TRUE),
                         grep("TJB\\d_MSUM", cols, value = TRUE, ignore.case = TRUE)),
              data.table = FALSE )

  # Rename columns to lowercase
  names(pu) <- tolower(names(pu))
  assign(newname, pu)
  saveRDS(newname, paste0("/Volumes/ExtremeSSD/SIPP Data Files/rds/",newname, ".rds"))
  
}








```

Here we impute those instances where someone had all of the earnings and job information we care about but did not have a jobid value for that specific job. This likely won't influence our models much since we don't have that many  we care about 
```{r}
library(tidyverse)
pu2014w1_reshaped_work_imputed <- pu2014w1 %>% 
  select('ssuid', 'spanel', 'swave', 'pnum', 'monthcode', contains("gamt"), contains("prftb"), contains("msum"), contains("mwkhrs"), contains("jborse"), contains("jobid"), ) %>% 
  tidylog::mutate(ejb1_jobid = ifelse(is.na(ejb1_jobid) & !is.na(tjb1_mwkhrs) & !is.na(ejb1_jborse) & (!is.na(tjb1_msum) | !is.na(tjb1_prftb)), 1, ejb1_jobid))
```

